\textbf{Seasonal Dynamics of Epiphytic Microbial Communities on Marine Macrophyte Surfaces}

### 2020-04-19 submission/manuscript.tex
References update.

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,750 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,408
(\autoref{supp-nseq_notus}), through rarefaction resulting in 16,972
different OTUs that ranged from 351 to 1,996 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,682.5 ±
134.4 OTUs{]} and monospecific {[}Number of OTUs, 1,727.4 ± 154.7
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,056 ±
201.8 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 523.8 ± 149.4 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.0 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (39.0 -- 51.6
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.54, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.30, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be found at every
sampling time (\(\leq\) 1.0 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 51.6
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems,
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass senescence. In contrast,
\emph{C. cylindracea} started to proliferate in late Spring and was
characterized only by a Summer community during maximal biomass increase
and by a Autumn/Winter/Spring community when the biomass was at the peak
and the settlement started to subsequently decay. Similar seasonal
changes in the epiphytic community was also described for other
macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011).
Higher temporal stability of \emph{C. cylindracea} surface communities
in comparison to \emph{C. nodosa} were also observed in the higher
proportion of shared communities between two consecutive sampling
points.

@@

\leavevmode\hypertarget{ref-Zavodnik1998}{}%
Zavodnik, N., Travizi, A., and De Rosa, S. (1998) Seasonal variations in
the rate of photosynthetic activity and chemical composition of the
seagrass \emph{Cymodocea nodosa} (Ucr.) Asch. \emph{Sci Mar}
\textbf{62}: 301--309.

>>>>

Sequencing of the 16S rRNA V4 region yielded a total of 1.8 million
sequences after quality curation and exclusion of eukaryotic,
chloroplast, mitochondrial and no relative sequences
(\autoref{supp-nseq_notus}). A total of 35 samples originating from
epiphytic archaeal and bacterial communities associated with surfaces of
the seagrass \emph{C. nodosa} and macroalga \emph{C. cylindracea} were
analyzed. In addition, 18 samples (one of the samples was sequenced two
times) originating from picoplankton archaeal and bacterial communities
in the ambient seawater were also processed for comparison. The number
of reads per sample ranged between 8,408 and 77,463 sequences
(\autoref{supp-nseq_notus}). Even when the highest sequencing effort was
applied the rarefaction curves did not level off that is a common
observation in high-throughput 16S rRNA amplicon sequencing approaches
(\autoref{supp-rarefaction}). Following quality curation and exclusion
of sequences mentioned before reads were clustered into 28,750 different
OTUs at a similarity level of 97 \si{\percent}. Read numbers were
normalized to the minimum number of sequences, 8,408
(\autoref{supp-nseq_notus}), through rarefaction resulting in 17,039
different OTUs that ranged from 380 to 2,023 OTUs per sample
(\autoref{supp-calculators}). To determine seasonal changes of richness
and diversity the Observed Number of OTUs, Chao1, ACE, Exponential
Shannon (Jost, 2006) and Inverse Simpson were calculated after
normalization through rarefaction. Generally, richness estimators and
diversity indices showed similar trends. On average, higher values were
found for \emph{C. cylindracea} (mixed {[}Number of OTUs, 1,687.6 ±
121.1 OTUs{]} and monospecific {[}Number of OTUs, 1,738.4 ± 160.6
OTUs{]}), middle values for \emph{C. nodosa} (Number of OTUs, 1,062 ±
218.2 OTUs) and lower values for picoplankton communities in the ambient
seawater (Number of OTUs, 524.8 ± 146.3 OTUs)
(\autoref{supp-calculators}). Seasonal changes did not show such large
dissimilarities. \emph{C. nodosa} communities showed a slow increase
towards the end of the study, while \emph{C. cylindracea} (mixed and
monospecific) communities were characterized by slightly larger values
in Spring and Summer in comparison to Autumn and Winter
(\autoref{supp-calculators}).

To determine the proportion of shared archaeal and bacterial OTUs and
communities sampled in different environments the Jaccard's Similarity
Coefficient on presence-absence data and Bray-Curtis Similarity
Coefficient were, respectively, calculated. Coefficients were determined
after normalization through rarefaction and binning of samples from a
particular environment. The highest proportion of shared OTUs and
community was found between mixed and monospecific \emph{C. cylindracea}
(Jaccard, 0.35; Bray-Curtis, 0.78), while lower shared values were
calculated between seawater and epiphytic communities
(\autoref{matrix}). Shared proportion between \emph{C. nodosa} and
\emph{C. cylindracea} were approximately in the middle between these two
extremes. To assess seasonal changes in the proportion of shared OTUs
and communities the Jaccard's and Bray-Curtis Similarity Coefficients
were calculated between consecutive sampling points (\autoref{shared}).
Both coefficients showed similar trends. Temporal proportional changes
were more pronounced for seawater in comparison to \emph{C. nodosa} and
especially \emph{C. cylindracea} associated communities
(\autoref{shared}). In addition, only 0.4 -- 1.1 \si{\percent} of OTUs
from each surface associated community were found at every time point.
These OTUs also made a high proportion of total sequences (39.6 -- 52.6
\si{\percent}). To further disentangle the environmental and seasonal
community dissimilarity a Principal Coordinates Analysis (PCoA) based on
Bray-Curtis distances and OTU abundances was applied. It showed a clear
separation between planktonic and surface associated communities
(\autoref{pcoa}). In addition, a separation of epiphytic bacterial and
archaeal communities based on host species was determined. This
separation was further supported by ANOSIM (R = 0.96, \emph{p}
\textless{} 0.001). Seasonal changes of \emph{C. nodosa} associated
communities indicated a separation between Spring, Summer and
Autumn/Winter samples (ANOSIM, R = 0.54, \emph{p} \textless{} 0.01). For
\emph{C. cylindracea} associated communities a separation between Summer
and Autumn/Winter/Spring samples was observed that was not so strongly
supported (ANOSIM, R = 0.32, \emph{p} \textless{} 0.01)
(\autoref{pcoa}).

@@

Seasonal richness changes in the epiphytic community were substantial as
could be observed in the proportion of OTUs that could be found at every
sampling time (\(\leq\) 1.1 \si{\percent}). Interestingly, these OTUs
were accounting for a high proportion of sequences (\(\leq\) 52.6
\si{\percent}). A very similar proportion of persistent OTUs and their
sequence contribution was reported in high frequency studies describing
seasonal picoplankton changes (Gilbert \emph{et al.}, 2009, 2012). In
comparison to the seawater community, a lower degree of seasonal shifts
was observed for the surface associated communities. It seems,
microniches on the surfaces of macrophytes are providing more stable
conditions in comparison to the seawater. At the level of OTUs seasonal
changes of \emph{C. nodosa} and \emph{C. cylindracea} associated
communities were identified that could be linked to the growth cycle of
the seagrass and macroalgae (M. Najdek, personal communication).
\emph{C. nodosa} was characterized by a Spring community during maximum
seagrass proliferation, a Summer community during a biomass maximum and
a Autumn/Winter community during a biomass senescence. In contrast,
\emph{C. cylindracea} started to proliferate in late Spring and was
characterized only by a Summer community during maximal biomass increase
and by a Autumn/Winter/Spring community when the biomass was at the peak
and the settlement started to subsequently decay. Similar seasonal
changes in the epiphytic community was also described for other
macroalgae (Tujula \emph{et al.}, 2010; Lachnit \emph{et al.}, 2011).
Higher temporal stability of \emph{C. cylindracea} surface communities
in comparison to \emph{C. nodosa} were also observed in the higher
proportion of shared communities between two consecutive sampling
points.

@@

\leavevmode\hypertarget{ref-Zavodnik1998}{}%
Zavodnik, N., Travizi, A., and de Rosa, S. (1998) Seasonal variations in
the rate of photosynthetic activity and chemical composition of the
seagrass \emph{Cymodocea nodosa} (Ucr.) Asch. \emph{Sci Mar}
\textbf{62}: 301--309.
